Isolator: accurate and consistent analysis of isoform-level expression in RNA-Seq experiments

### 2016-02-16 paper.tex
Add Nathan's story. Misc improvements.

\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{sectsty}
\usepackage{ctable}
\usepackage{hyperref}
\usepackage{subcaption}
\usepackage[numbers]{natbib}
\usepackage{xr}

@@

\title{Isolator: accurate and consistent analysis of RNA-Seq experiments}

% Possible alternative titles? Larry suggests putting 'isoform' in the name.
% 
% accurate and consistent isoform quantification
%
%
%

\author{Daniel C. Jones \and Kavitha T. Kuppusamy \and Xinxia Peng \and Michael G. Katze \and Hannele Ruohola-Baker \and Walter L. Ruzzo}
\date{}

@@

\begin{abstract}
    While RNA-Seq has enabled great progress towards the goal of wide-scale
    isoform-level mRNA quantification, short reads have limitations when
    resolving complex or similar sets of isoforms. As a result, estimates of
    isoform abundance carry far more uncertainty than those made at the gene
    level. When confronted with this uncertainty, commonly used methods produce
    estimates that are unstable; small perturbations in the data often produce
    dramatically different results confounding downstream analysis. We introduce
    a new method, Isolator, which analyzes all samples in an experiment in
    unison using a simple Bayesian hierarchical model. Combined with aggressive
    bias correction, it produces estimates that are simultaneously accurate and
    stable. In a comprehensive comparison of accuracy and stability, we show
    that this property is unique to Isolator. We further demonstrate that the
    approach of modeling an entire experiment enables new analyses by examining
    splicing monotonicity across several time points in the development of human
    cardiomyocyte cells.
\end{abstract}

@@

Since its introduction, RNA-Seq has rapidly become a preferred method of
studying gene expression, having proved to be reliable, reproducible, and
increasingly affordable. In principle, RNA-Seq enables measurements of
expression at a scale and resolution than has previously been impossible or
impractical. Rather than relying on an often vaguely defined notion of a gene,
expression of specific isoforms or exons can be measured. In practice this
promise is difficult to realize. There are no standardized names or symbols for
isoforms, annotations are often incomplete or in disagreement, and methods
designed to detect changes in splicing rarely concur \cite{Liu:2014dn}.

Most RNA-Seq procotocols produce short reads (typically 30 to 150 nucleotides).
Because a short read is often compatible with many isoforms, the signal produced
from sequencing must be deconvoluted, requiring a certain level of algorithmic
sophistication. This problem was tackled early on and with great success by
Cufflinks \cite{Trapnell:2010kd}, and indeed many of the methods developed in
the interim have followed from the same basic model of RNA-Seq in which
transcripts are represented as particular distributions over possible reads, and
the problem is then to infer mixing coefficients (relative expression values)
inducing a mixture model that best explains the data \cite{Pachter:2011wm}.

We built on this work, implementing a new method called Isolator. Though
it retains the same underlying model of RNA-Seq, we approach the problem
from the perspective that isoform-level estimates possess a degree
uncertainty that renders maximum likelihood point-estimates alone a
fundamentally inadequate solution to the problem. As the purpose of gene
expression studies is to compare expression between multiple samples,
our goal, beyond maximizing a narrow definition of accuracy, is to
provide a tool that effectively accounts for this uncertainty in a
coherent and reliable manner.

Rather than estimating expression of individual samples, Isolator uses a
hierarchical model of an entire experiment. In addition to inferring
expression values for the individual samples, we introduce parameters
representing condition-wise and experiment-wise expression and splicing,
as well as per-transcript variance parameters that are shared across
conditions. As expression studies often use only a small number of
replicates, shared variance parameters allows us to make more accurate
estimates of biological variability.

An efficient Markov chain Monte Carlo algorithm is used to generate samples over
the parameters of this model, which are saved in an HDF5 \cite{hdf5} based
format. The output from the sampler can then be processed to produce point
estimates, credible intervals, posterior probabilities, and diagnostics. Though
designed to be run on multiple samples concurrently, Isolator can be run as a
conventional tool on individual samples.

The broad effect of this model is to encode the common assumption that
genes tend to be similarly expressed and similarly spliced between
replicates, and to a lesser extent between conditions. In the absence of
sufficient evidence to the contrary, this informative prior shrinks
estimates towards a baseline of no change, producing more conservative
estimates of effect size than would be obtained by using, for example, a
flat Dirichlet prior. This feature of the model is particularly
important for robust treatment of low-expression transcripts. With few reads
there is little power to determine splicing patterns. Considering the
samples in isolation will often produce very different estimates.
Statistical tests based on these estimates, without considering the full
posterior distribution, make approximations by either either
disregarding the relative or compositional nature of the data and
assuming independence, or by relying on asymptotic properties that may
not hold when a gene is not deeply sequenced.

In addition to this hierarchical model, we attempt to aggressively correct for
technical effects. The naive model of RNA-Seq assumes fragments are sampled
uniformly at random from transcripts in proportion to their abundance. Several
have observed that this uniformity assumption is far from an accurate
description of real sequencing data \cite{Hansen:2010if}. Though severe at the
nucleotide resolution, gene level expression estimates are only moderately
affected by this bias, as it tends to average out over long transcripts. When
one considers expression at the isoform level, attempting to correct for this
nonuniformity becomes vital, as the difference between two isoforms is often
only a few nucleotides.

We previously developed an efficient and versatile model to account for sequence
specific bias that commonly occurs at the ends of fragments \cite{Jones:2012hg}
In Isolator, we supplemented this with methods accounting for 3' bias, fragment
GC-content bias, and positional bias caused by end-effects of fragmentation.

@@

The convention for evaluating the accuracy of RNA-Seq analysis methods has
consisted chiefly of comparing point estimate accuracy in simulations,
supplemented with comparisons to qPCR or other technologies. Tuning methods
exclusively to maximize accuracy on simulated data risks solving only an
idealized mathematical model of RNA-Seq while disregarding noise, bias, and
nonuniformity. As no gold standard for whole-transcriptome isoform-level
estimation exists, we adopt multi-faceted approach, with a focus on measuring
the stability or consistency of estimates, along with accuracy.

Comparisons of gene or transcript expression are sensitive to how one
chooses to measure similarity or distance. Because gene expression
typically varies across many orders of magnitude, measurements like
the Pearson correlation, $L_1$ or $L_2$ distance, or root-mean-square
error are often dominated by a small subset of the most highly expressed
genes. Recognizing this, a popular alternative is the Spearman rank
correlation, which conversely can overemphasize the majority of
transcripts which have no, or very low expression. These issues are
discussed in depth by Lovell et al \cite{Lovell:2015il}, who propose
measurements of proportionality as a more meaningful metric. We adopt
their ``proportionality correlation'', defined as
$\text{pcor}(x, y) = 2\text{cov}(\log(x), \log(y)) / (\text{var}(\log(x)) + \text{var}(\log(y)))$.
Similar to other measures of correlation, proportionality correlation
varies from -1 to 1, with 1 indicating perfectly proportionality, and -1
perfect reciprocality. Zeros are accounted for by converting all measurements to
transcripts per million (TPM) and adding 0.1.

@@

RSEM was used to produce both maximum likelihood and posterior
mean estimates, which we label ``RSEM/ML'' and ``RSEM/PM'', respectively.
This is not an exhaustive account of such methods, but does represent a
wide variety of popular approaches.

@@

We use data generated by the Sequencing Quality Control Consortium (SEQC)
\cite{Consortium:2014kc}, consisting of four reference samples (labeled A, B, C,
and D). The RNA-seq methods were run using approximately 25 million reads per
sample, obtaining estimates that were then compared to qPCR in Figures
\ref{table:taqman} and \ref{table:primepcr}, and the known ERCC spike-in
mixtures in Figure \ref{table:ercc}. In each of these we find that Isolator
produces the highest correlation of the methods compared, though sometimes by a
negligible margin. Despite two different technologies, and two different sets of
genes, the qPCR benchmarks agree closely, ranking the methods nearly
identically. Cufflinks, eXpress, Salmon, and RSEM/ML typically perform similarly
to each other, ith BitSeq, Kallisto, Sailfish, and RSEM/PM trailing slightly.
Results are similar with ERCC spike-in controls, the notable exceptions being
Kallisto and Sailfish, which show higher accuracy, and eXpress, which shows
lower accuracy compared to their relative rankings in the qPCr benchmarks.

@@

Results from RSEM are particularly informative. We generated maximum
likelihood and posterior mean estimates using the same aligned reads.
Merely switching the estimate from the former to the latter increased
the correlation from 0.922 to 0.968. This stands in contrast to
comparisons to qPCR which suggest reduced accuracy of posterior mean
estimates, demonstrating that consistency and accuracy, while related,
are two separate axis of comparison.

@@

\begin{figure*}
\includegraphics[width=\textwidth]{batch-effect-figure.pdf}
\caption{
\textbf{a} A heatmap showing pairwise proportionality correlation between the
samples sequenced on two flowcells each at five sites. Flowcells are numbered 1
or 2 and sequencing sites are abbreviated with three letter codes: Australian
Genome Research Facility (AGR), Beijing Genome Institute (BGI), Cornell
University (CNL), Mayo Clinic (MAY), and Novartis (NVS). \textbf{b} The absolute
change in correlation induced by enabling bias correction for methods that
support it. For clarity this plot excludes points for BitSeq estimates of "MAY
2", as bias correction has an extreme detrimental effect on these.}
\label{fig:batcheffects}
\end{figure*}

@@

These correlations are somewhat smaller than those in Table
\ref{table:seqctranscripts},
largely because of shallower sequencing: there five flowcell lanes were
combined, totaling \textasciitilde{}25 millions reads per sample, here
only one lane or \textasciitilde{}5 million reads are used.
Nevertheless, we again see posterior mean estimates with significantly higher
agreement between pairs.

When accounting for batch effects, bias correction methods can have a
significant impact. To examine the efficacy of bias correction, we disabled this
functionality on those methods that support it, repeated the experiment, and
measured the change in pairwise correlation (Figure \ref{fig:batcheffects}b). We
see that bias correction is largely beneficial for each method. Isolator,
Salmon, and Cufflinks show similar improvements with bias correction, though in
a few cases Cufflinks slightly decreases agreement. Kallisto show a consistent,
but very slight improvement. BitSeq's bias correction sometimes has a very positive
effect, but other times negative or disastrous effect. Estimates from the ``MAY
2'' samples in particular had far worse agreement with other samples with bias
correction enabled. Here bias correction has dramatic and detrimental effects on
some number of transcripts, driving the correlation to negative numbers.

While this benchmark is informative, it should be considered a lower bound on
the batch effects and bias: these samples were identically prepared and
sequenced. The fact that technical effects were so clearly present even under
such controlled circumstances is indicative that RNA-Seq is not so reproducible
that batch effects can be, under any circumstanced ignored.

@@

To more directly demonstrate Isolator's capacity to produce more accurate
estimates by modeling entire experiments, we simulated a simple RNA-Seq
experiment consisting of two conditions each with three replicated. Expression
values were generated from a two-component log-normal model with parameters fit
to Cufflinks estimates of data from Kuppusamy et al \cite{Kuppusamy:2015ey}. We
generated 10 million, 100nt paired-end simulated RNA-Seq reads per sample using
rlsim \cite{Sipos:2013tz}.

In this benchmark, Isolator significantly improves on existing methods.
It has the unique advantage of sharing information between replicates.
With the assumption that genes tend to be similarly spliced between
replicates, we are able to more effectively resolve transcript
expression in complex loci. Kallisto, Salmon, RSEM/ML, and Cufflinks all
perform very similarly. While comparisons to qPCR showed eXpress and
Cufflinks with a significant advantage, effective bias correction is
less important in simulated data, plausibly explaining the difference.
Although rlsim models some forms of bias, technical effects in RNA-Seq
are not entirely understood, so these are likely understated in its model.

@@

Isolator shows the highest correlation, but only when sequencing depth
exceeds several million reads, where it overtakes eXpress by a small
margin. The outliers in the this test are BitSeq and RSEM/PM. The former
shows extremely high correlation with small numbers of reads, but
contrary to all other methods, correlation rapidly decreases with deeper
sequencing after 1 million reads. When posterior mean estimates are
generated from a model using a uninformative prior, as is the case with
BitSeq, transcripts with few or no reads may produce estimates that are
highly influenced by transcript length, which plausibly explains this
phenomenon. Yet, RSEM/PM does increase in correlation monotonically with
sequencing depth, though the correlation remains very low. Even using
all 11 million reads, RSEM/PM had a correlation of only 0.567.

@@

TODO: add nathan's text here

>>>>

\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{sectsty}
\usepackage{ctable}
\usepackage{hyperref}
\usepackage{subcaption}
\usepackage[numbers]{natbib}
\usepackage{xr}
\usepackage{authblk}

@@

\title{Isolator: accurate and consistent analysis of isoform-level expression in RNA-Seq experiments}

\author[1]{Daniel C. Jones}
\author[2]{Kavitha T. Kuppusamy}
\author[3]{Nathan J. Palpant}
\author[4]{Xinxia Peng}
\author[4]{Michael G. Katze}
\author[5]{Charles E. Murry}
\author[6]{Hannele Ruohola-Baker}
\author[1,7,8]{Walter L. Ruzzo}

%\date{}

\affil[1]{Department of Computer Science and Engineering, University of Washington}
\affil[2]{XXX}
\affil[3]{XXX}
\affil[4]{Department of Microbiology, University of Washington}
\affil[5]{Department of Pathology, University of Washington}
\affil[6]{Department of Biochemistry, University of Washington}
\affil[7]{Department of Genome Sciences, University of Washington}
\affil[8]{Fred Hutchinson Cancer Research Center}

%\affil[1]{Department of Computer Science and Engineering,
%University of Washington, Seattle, WA 98195-2350, USA}
%\affil[2]{Department of Genome Sciences, University of Washington, Seattle, WA
%98195-5065, USA}
%\affil[3]{Fred Hutchinson Cancer Research Center, Seattle, WA 98109, USA}
%\affil[4]{Department of Microbiology, University of Washington, Seattle, WA
%98195-7242, USA}

@@

\begin{abstract}
    While RNA-Seq has enabled great progress towards the goal of wide-scale
    isoform-level mRNA quantification, short reads have limitations when
    resolving complex or similar sets of isoforms. As a result, estimates of
    isoform abundance carry far more uncertainty than those made at the gene
    level. When confronted with this uncertainty, commonly used methods produce
    estimates that are unstable---small perturbations in the data often produce
    dramatically different results, confounding downstream analysis. We
    introduce a new method, Isolator, which analyzes all samples in an
    experiment in unison using a simple Bayesian hierarchical model. Combined
    with aggressive bias correction, it produces estimates that are
    simultaneously accurate and stable. In a comprehensive comparison of
    accuracy and stability, we show that this property is unique to Isolator. We
    further demonstrate that the approach of modeling an entire experiment
    enables new analyses, which we demonstrate by examining splicing
    monotonicity across several time points in the development of human
    cardiomyocyte cells.
\end{abstract}

@@

Since its introduction, RNA-Seq has rapidly become a preferred method of
studying gene expression, having proved to be reliable, reproducible, and
increasingly affordable. In principle, RNA-Seq enables measurements of
expression at a scale and resolution than has previously been impractical or
impossible. Rather than relying on an often vaguely defined notion of a gene,
expression of specific isoforms or exons can be measured. In practice this
promise is difficult to realize, with methods designed to detect changes in
splicing rarely concuring \cite{Liu:2014dn}.

Most RNA-Seq procotocols produce short reads (typically 30 to 150 nucleotides,
often paired-end).  Because a short read is often compatible with many isoforms,
the signal produced from sequencing must be deconvoluted, requiring a certain
level of algorithmic sophistication. This problem was tackled early on and with
great success by Cufflinks \cite{Trapnell:2010kd}, and indeed many of the
methods developed in the interim have followed from the same basic model of
RNA-Seq in which transcripts are represented as particular distributions over
possible reads, and the problem is then to infer mixing coefficients (relative
expression values) inducing a mixture model that best explains the data
\cite{Pachter:2011wm}.

We built on this work, implementing a new method called Isolator. Though
it retains the same underlying model of RNA-Seq, we approach the problem
from the perspective that isoform-level estimates possess a degree of
uncertainty that renders maximum likelihood point-estimates alone a
fundamentally inadequate solution to the problem. As the purpose of gene
expression studies is to compare expression between multiple samples,
our goal, beyond maximizing a narrow definition of accuracy, is to
provide a tool that effectively accounts for this uncertainty in a
coherent and reliable manner.

Rather than estimating expression of individual samples, Isolator uses a
hierarchical model of an entire experiment, for example, including all samples
and time points in a time series. In addition to inferring expression values for
the individual samples, we introduce parameters representing condition-wise and
experiment-wise expression and splicing, as well as per-transcript variance
parameters that are shared across conditions. As expression studies often use
only a small number of replicates, shared variance parameters allows us to make
more accurate estimates of biological variability.

An efficient Markov chain Monte Carlo algorithm is used to generate samples over
the parameters of this model, which are saved in an HDF5-based format
\cite{hdf5}. The output from the sampler can then be processed to produce point
estimates, credible intervals, posterior probabilities, and diagnostics. Though
designed to be run on multiple samples concurrently, Isolator can be run as a
conventional tool on individual samples.

The broad effect of this model is to encode the common assumption that
genes tend to be similarly expressed and similarly spliced between
replicates, and to a lesser extent between conditions. In the absence of
sufficient evidence to the contrary, this informative prior shrinks
estimates towards a baseline of no change, producing more conservative
estimates of effect size than would be obtained by using, for example, a
flat Dirichlet prior. This feature of the model is particularly
important for robust treatment of low-expression transcripts. With few reads
there is little power to determine splicing patterns. Considering the
samples in isolation will often produce very different estimates.
Statistical tests based on these estimates, without considering the full
posterior distribution, make approximations by either
disregarding the relative or compositional nature of the data and
assuming independence, or by relying on asymptotic properties that may
not hold when a gene is not deeply sequenced.

In addition to this hierarchical model, we attempt to aggressively correct for
technical effects. The naive model of RNA-Seq assumes fragments are sampled
uniformly at random from transcripts in proportion to their abundance. Early on,
this was shown to be far from an accurate description of real sequencing data
\cite{Hansen:2010if}. Though severe at nucleotide resolution, gene level
expression estimates are only moderately affected by this bias, as it tends to
average out over long transcripts. When one considers expression at the isoform
level, attempting to correct for this nonuniformity becomes vital, as the
difference between two isoforms is often only a few nucleotides.

We previously developed an efficient and versatile model to account for sequence
specific bias that commonly occurs at the ends of fragments \cite{Jones:2012hg}.
In Isolator, we supplemented this with methods accounting for 3' bias, fragment
GC-content bias, and positional bias caused by end-effects of fragmentation.

@@

The convention for evaluating the accuracy of RNA-Seq analysis methods has
consisted chiefly of comparing point estimate accuracy in simulations,
supplemented with comparisons to qPCR or other technologies. Tuning methods
exclusively to maximize accuracy on simulated data risks solving only an
idealized mathematical model of RNA-Seq while disregarding noise, bias, and
nonuniformity. As no gold standard for whole-transcriptome isoform-level
estimation exists, we adopt a multifaceted approach, focusing on measuring
the stability or consistency of estimates, along with accuracy.

Comparisons of gene or transcript expression are sensitive to how one chooses to
measure similarity. Because gene expression typically varies across many orders
of magnitude, measurements like the Pearson correlation, $L_1$ or $L_2$
distance, or root-mean-square error are often dominated by the most highly
expressed genes. Recognizing this, a popular alternative is Spearman's rank
correlation, which conversely can overemphasize the majority of transcripts,
which have no, or very low expression. These issues are discussed in depth by
Lovell et al \cite{Lovell:2015il}, who propose measurements of proportionality
as a more meaningful metric. We adopt their ``proportionality correlation'',
defined as $\text{pcor}(x, y) = 2\text{cov}(\log(x), \log(y)) /
(\text{var}(\log(x)) + \text{var}(\log(y)))$, where is $\text{cov}(x)$ and
$\text{var}(x)$ give the covariance and variance of $x$, respectively.  Similar
to other measures of correlation, proportionality correlation varies from -1 to
1, with 1 indicating perfectly proportionality, and -1 perfect reciprocality.
Zeros are accounted for by converting all measurements to transcripts per
million (TPM) and adding 0.1.
% TODO: add supplementary info about why 0.1 is a reasonable value.

@@

RSEM was used to produce both maximum likelihood and posterior mean estimates,
which we label ``RSEM/ML'' and ``RSEM/PM'', respectively.  This is not an
exhaustive account of such methods, but does represent a wide variety of popular
approaches. When possible, the same read alignments were used for each method,
which notably excludes the alignment-free methods Sailfish, Salmon, and
Kallisto.
% TODO: see supplementary section XXX

@@

We use data generated by the Sequencing Quality Control Consortium (SEQC)
\cite{Consortium:2014kc}, consisting of four reference samples (labeled A, B, C,
and D). The RNA-seq methods were run using approximately 25 million reads per
sample, obtaining estimates that were then compared to qPCR in Figures
\ref{table:taqman} and \ref{table:primepcr}, and the known ERCC spike-in
mixtures in Figure \ref{table:ercc}. In each of these we find that Isolator
produces the highest correlation of the methods compared, though sometimes by a
small margin. Despite two different technologies, and two different sets of
genes, the qPCR benchmarks agree closely, ranking the methods nearly
identically. Cufflinks, eXpress, Salmon, and RSEM/ML typically perform similarly
to each other, with BitSeq, Kallisto, Sailfish, and RSEM/PM trailing slightly.
Results are similar with ERCC spike-in controls, the notable exceptions being
Kallisto and Sailfish, which show higher accuracy, and eXpress, which shows
lower accuracy compared to their relative rankings in the qPCR benchmarks.

@@

Results from RSEM are particularly informative. We generated maximum likelihood
and posterior mean estimates using the same aligned reads. Merely switching the
estimate from the former to the latter increased the correlation by
approximately 5\%. This stands in contrast to comparisons to qPCR which suggest
reduced accuracy of posterior mean estimates, demonstrating that consistency and
accuracy, while related, are two separate axis of comparison.

@@

\begin{figure*}
\includegraphics[width=\textwidth]{batch-effect-figure.pdf}
\caption{
\textbf{a} A heatmap showing pairwise proportionality correlation between
samples sequenced on two flowcells each at five sites, from centrally prepared
libraries. Flowcells are numbered arbitrarily 1 or 2 and sequencing sites are
abbreviated with three letter codes: Australian Genome Research Facility (AGR),
Beijing Genome Institute (BGI), Cornell University (CNL), Mayo Clinic (MAY), and
Novartis (NVS). \textbf{b} The absolute change in correlation induced by
enabling bias correction for methods that support it. For clarity this plot
excludes points for BitSeq estimates of "MAY 2", as bias correction has an
extremely detrimental effect on these. Mean improvement in correlation was 0.008
with Salmon, 0.007 with Cufflinks, 0.006 with Isolator, 0.003 with Kallisto,
0.002 with eXpress, and -0.164 with BitSeq.
}

\label{fig:batcheffects}
\end{figure*}

@@

These correlations are somewhat smaller than those in Table
\ref{table:seqctranscripts}, largely because of shallower sequencing: in Table
\ref{table:seqctranscripts} five flowcell lanes were combined, totaling
\textasciitilde{}25 millions reads per sample, here only one lane
(\textasciitilde{}5 million reads) was used. Nevertheless, we again see
posterior mean estimates (as in Isolator, BitSeq, and RSEM/PM) with
significantly higher agreement between pairs.

When accounting for batch effects, bias correction methods can have a
significant impact. To examine the efficacy of bias correction, we disabled this
functionality on those methods that support it, repeated the experiment, and
measured the change in pairwise correlation (Figure \ref{fig:batcheffects}b). We
see that bias correction is largely beneficial for each method. Isolator,
Salmon, and Cufflinks show similar improvements with bias correction, though in
a few cases Cufflinks slightly decreases agreement. Kallisto show a consistent,
but very slight improvement. BitSeq's bias correction sometimes has a very
positive effect, but other times a negative or even disastrous effect. Estimates
from the ``MAY 2'' samples in particular had far worse agreement with other
samples with bias correction enabled. Here bias correction has dramatic and
detrimental effects on some number of transcripts, driving the correlation to
negative numbers.

While this benchmark is informative, it should be considered a lower bound on
the batch effects and bias found in real experiments: these samples were
sequenced at different labs, but using the same instrument model (HiSeq 2000)
and from libraries prepared centrally. The fact that technical effects were so
clearly present even under such controlled circumstances suggests that
reproducibility should not be taken for granted in RNA-Seq experiments,
particularly when it comes to batch effects.

@@

To more directly demonstrate Isolator's capacity to produce more accurate
estimates by modeling entire experiments, we simulated a simple RNA-Seq
experiment consisting of two conditions each with three replicates. Expression
values were generated from a two-component log-normal model with parameters fit
to Cufflinks estimates of data from Kuppusamy et al \cite{Kuppusamy:2015ey}. We
generated 10 million, 100nt paired-end simulated RNA-Seq reads per sample using
rlsim \cite{Sipos:2013tz}.

In this benchmark, Isolator significantly improves on existing methods (Table
\ref{table:rlsim}). It has the unique advantage of sharing information between
replicates. With the assumption that genes tend to be similarly spliced between
replicates, we are able to more effectively resolve transcript expression in
complex loci. Kallisto, Salmon, RSEM/ML, and Cufflinks all perform very
similarly. While comparisons to qPCR showed eXpress and Cufflinks with a
significant advantage over RSEM/ML, Kallisto, and Bitseq, effective bias
correction is less important in simulated data, plausibly explaining the
difference. Although rlsim models some forms of bias, technical effects in
RNA-Seq are not entirely understood, so these are likely understated in its
model.

@@

Isolator shows the highest correlation, but only when sequencing depth
exceeds a few million reads (below which would be unusually shallow
sequencing for an RNA-Seq experiment) where it overtakes eXpress by a small
margin. The outliers in the this test are BitSeq and RSEM/PM. The former
shows extremely high correlation with small numbers of reads, but
contrary to all other methods, correlation rapidly decreases with deeper
sequencing after 1 million reads. When posterior mean estimates are
generated from a model using a uninformative prior, as is the case with
BitSeq, transcripts with few or no reads may produce estimates that are
highly influenced by transcript length, which plausibly explains this
phenomenon. Yet, RSEM/PM does increase in correlation monotonically with
sequencing depth, though the correlation remains very low. Even using
all 11 million reads, RSEM/PM had a correlation of only 0.567.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=0.75\textwidth]{palpant/figure.pdf}
\end{center}
\caption{
\textbf{a}
Hierarchical clustering of the dominant splice variant for 84
differentially spliced genes. Blue arrows indicate genes with known function in
endothelial or heart development. Red arrows indicate genes in the class of
receptors that bind VEGF.
\textbf{b}
Two dimensional principle component analysis in mesodermal progenitor
populations showing discrete separation of populations entirely based on
differences in splice variants. 
\textbf{c}
Analysis of gene ontology showing enriched categories found in 84 differentially
spliced genes. (d-e) Heat map analysis 
\textbf{d}
Heat map showing expression of annotated isoforms of NRP1.
\textbf{e}
Read coverage over isoforms of NRP1.
\textbf{f}
Schematic diagram of protein domains encoded by splice variant variants of NRP1.}
\label{fig:mesodermal}
\end{figure*}

Very little is known about the role of splice variants in mediating cell fate
specification even though it has been shown that 92-95\% of genes are
differentially spliced in a cell or tissue specific manner \cite{Wang:2008ea,
Pan:2008jq}. In order to identify novel mechanisms of cell fate specification in
mesoderm development we began by analyzing RNA-seq data to determine differences
in splice variants between human embryonic stem cell (hESC) derived cardiac
progenitor cells (CPCs) and two endothelial lineages, hESC-derived endocardial
endothelial cells (EECs) and hemogenic endothelial cells (HECs) as we have
described previously \cite{Palpant:2015dv}. 84 genes were identified as
differentially spliced (defined here as a > 0.3 shift in relative transcript
abundance with > 0.5 posterior probability) between these populations (Figure
\ref{fig:mesodermal}a).  Principle component analysis indicated that splice
variants, independent of gene expression levels, delineate endothelial
populations relative to the CPCs (Figure \ref{fig:mesodermal}b). Furthermore,
these data indicate that splice variants alone are sufficient to differentiate
these closely related lineages.  Gene ontology analysis of the 84 differentially
spliced genes showed appropriate enrichment for categories associated with
alternative splicing (Figure \ref{fig:mesodermal}c). Upon closer examination, we
identified neuropilin-1 (NRP1) as an excellent candidate molecule that showed
highly distinct expression profiles of splice variants between the cardiac and
hemogenic endothelial lineages (Figure \ref{fig:mesodermal}d-f).

NRP1 plays a versatile set of roles involving regulation of signaling pathways
that mediate angiogenesis, axon guidance, cell survival, migration, and cell
invasion in a wide range of developmental states and tissue types
\cite{Zachary:2011dr}. It is a single pass transmembrane molecule that has
multiple extracellular functional domains that interact with VEGF and/or
semaphorin molecules \cite{Cackowski:2004ja}. NRP1 has been implicated in
endothelial \cite{Lee:2002dh} and cardiac \cite{Gu:2003vq} development but the
specific splice variants involved in mediating these processes has not been
elucidated.  Results shown here indicate that the cardiac splice variant
contains all functional domains involved in binding VEGF and semaphorin 3a
(Figure \ref{fig:mesodermal}d-f). This is consistent with studies showing the
requirement for NRP1 in mediating the activities of VEGF and semaphorins in
cardiac development \cite{Gu:2003vq}. In contrast, we show that splice variants
expressed in endothelial populations include truncated versions that either lack
the semaphorin binding CUB domains and, in the case of HECs, express a secreted
version of NRP1 (Figure \ref{fig:mesodermal}d-f). This suggests that VEGF and
not semaphorin 3a is the primary mechanism of signaling through NRP1 in
endothelial development \cite{Lee:2002dh}. These results provide the first
evidence to our knowledge that NRP1 splice variants may be key determinants of
fate determination in mesoderm patterning.
